\section{Conclusions}
\label{sec:conclusions}

We embedded the Bodnar et al.~\cite{bodnar2016direct} direct precision shrinkage framework inside the dual Ensemble Kalman Filter, recomputing the shrinkage coefficients at each assimilation cycle, and benchmarked six estimators on two systems with opposite precision structure. Three findings emerge.

First, target-truth alignment determines direct precision shrinkage performance. With near-diagonal cosmological data, all four diagonal targets we tested yield well-behaved coefficients and the methods achieve the lowest standalone Frobenius loss. With dense-precision Lorenz~96 data, the same methods produce coefficients that require clamping in more than 80\% of cycles, $\beta^*$ values that drift to $10^8$--$10^9$, and condition numbers of $10^7$--$10^{11}$. The failure is a direct consequence of matching a diagonal target to a dense truth, not a flaw in the Bodnar et al.~\cite{bodnar2016direct} framework.

Second, good standalone estimation does not imply good filtering. Ledoit--Wolf produces the worst standalone precision estimates on cosmological data (Frobenius loss $\approx 0.93$--$0.96$ versus $0.41$--$0.46$ for the direct precision methods) yet achieves the lowest EnKF RMSE at every tested ensemble size. Ledoit--Wolf's cycle-to-cycle variance is an order of magnitude below the direct precision methods', and over 100 sequential assimilation cycles that stability outweighs the bias penalty. For sequential data assimilation, variance reduction is at least as important as bias reduction.

Third, the scaled identity target we propose, $\boldsymbol{\Pi}_0 = \mathrm{diag}(1/(2\sigma_i^2))$, fills a middle ground between the plain identity (no scale information) and the cosmological target (requires domain knowledge). It gives modest improvements over the plain identity target on both test systems without requiring any domain-specific information, and serves as a reasonable default for near-diagonal precision when no domain-specific target is available.

Practical guidance for choosing an EnKF precision estimator follows from these findings: in the absence of a domain-specific target that matches the true precision structure, Ledoit--Wolf is the safest default by virtue of its low cycle-to-cycle variance; direct precision shrinkage is preferable when a matched target exists \emph{and} standalone estimation quality is the objective; NERCOME is ruled out for sequential use by its $\mathcal{O}(B)$ cost; and the modified Cholesky decomposition remains the only tractable option at operational state dimensions. Future work includes a hybrid that combines localization with direct precision shrinkage, an adaptive target selection mechanism that diagnoses the off-diagonal structure of the sample precision, and a theoretical treatment of the sequential accumulation effect via a Lyapunov analysis of the filter error. All implementations are publicly available in the TEDA toolbox\footnote{\url{https://github.com/enino84/TEDA.git}}.

\subsubsection{Acknowledgements.} This work was supported by the Department of Computer Science and Statistics at Universidad del Norte.
