\section{Discussion}
\label{sec:discussion}

\subsection{Numerical Behavior of Shrinkage Coefficients}

The closed-form coefficients $\alpha^*$ and $\beta^*$ from Bodnar et al.~\cite{bodnar2016direct} behave poorly on Lorenz~96. While $\alpha^*$ stays near $1 - n/N_e$ across all ensemble sizes and assimilation steps, $\beta^*$ is severely miscalibrated: unclamped values reach $10^8$--$10^9$ for the identity target and $10^3$--$10^5$ for the scaled target, requiring clamping to $[0,1]$ at every single assimilation step. For the eigenvalue target, $\beta^*$ is effectively zero ($\sim 10^{-4}$), so no shrinkage toward the target occurs at all.

This happens because the targets are far from the true precision structure of Lorenz~96, which has significant off-diagonal entries reflecting spatial coupling in the dynamics. The identity target $\boldsymbol{\Pi}_0 = \mathbf{I}$ is a particularly poor match. In the Bodnar formula for $\beta^*$ (Eq.~\ref{eq:beta_star}), the $\text{tr}(\hat{\boldsymbol{\Pi}}_S \boldsymbol{\Pi}_0) / \|\boldsymbol{\Pi}_0\|_F^2$ factor becomes very large when the sample precision has large diagonal entries relative to the target norm, driving $\beta^*$ to extreme values.

After clamping, the estimator reduces to approximately $\boldsymbol{\Pi}_{\text{LS}} \approx \alpha^* \hat{\boldsymbol{\Pi}}_S + (1 - \alpha^*) \boldsymbol{\Pi}_0$ for the identity and scaled targets, or just $\boldsymbol{\Pi}_{\text{LS}} \approx \alpha^* \hat{\boldsymbol{\Pi}}_S$ for the eigenvalue target. The resulting precision matrices have condition numbers of $10^7$--$10^{11}$, producing analysis updates that are numerically unstable and highly sensitive to the random seed.

\subsection{Effect of Shrinkage Target Choice}

Covariance-domain and precision-domain shrinkage behave quite differently in our experiments. Ledoit-Wolf operates in covariance space with an analytically optimal shrinkage intensity that adapts to the data, achieving the lowest and most stable RMSE across all ensemble sizes on Lorenz~96.

The direct precision shrinkage methods, on the other hand, show erratic non-monotonic behavior as $N_e$ varies---RMSE can jump at certain ensemble sizes before dropping again. This is not a bug but a consequence of the Bodnar formula producing coefficients outside their valid range, which clamping cannot fully fix. These methods were developed and validated for cosmological precision matrices~\cite{looijmans2024comparison}, where targets like $\boldsymbol{\Pi}_0^{(2)}$ match the true precision structure. Applied to a system with different covariance characteristics, the target mismatch degrades performance.

Modified Cholesky gives stable (though not optimal) performance across all ensemble sizes, making it a reliable baseline.

Our scaled identity target occupies a middle ground. On Lorenz~96, it outperforms the other direct precision variants (Table~\ref{tab:performance}), probably because scaling by empirical variances gives a better-conditioned target than the plain identity. On cosmological data, it achieves the lowest mean Frobenius loss at $n = 30$ (Table~\ref{tab:cosmo_direct}) but falls slightly behind the Bodnar methods at larger sample sizes. It serves its intended purpose as a domain-agnostic alternative: modest gains over the identity target without requiring external information, but no substitute for a domain-specific target when one is available.

NERCOME shows a counterintuitive trend on Lorenz~96: RMSE \emph{increases} with ensemble size ($0.23$ at $N_e = 50$ to $0.42$ at $N_e = 100$). We attribute this to the interaction between NERCOME's bootstrap resampling (200 draws per assimilation cycle) and the sequential nature of the EnKF. At small $N_e$, the sample covariance is noisy and the resampling provides genuine regularization; at larger $N_e$, the sample covariance is already reasonable, and bootstrap variance compounds over 100 sequential updates. This effect does not appear on the cosmological data (Table~\ref{tab:cosmo_enkf}), where NERCOME's performance is stable. NERCOME also has a significant computational cost: on the AR(1) model ($d = 18$, $N_e = 30$), it takes ${\sim}27$~seconds per 100 cycles versus ${\sim}0.2$~seconds for the other shrinkage methods and ${\sim}0.14$~seconds for the standard EnKF. On Lorenz~96 ($n = 40$, $N_e = 80$), the gap widens to ${\sim}215$~seconds versus ${\sim}26$~seconds, an $8\times$ overhead from the bootstrap resampling.

\subsection{Regime Dependence on $N_e / n$}

For the standard EnKF, larger ensembles monotonically improve performance. Ledoit-Wolf is nearly flat across the tested range ($N_e/n \in [1.25, 2.5]$) thanks to its adaptive shrinkage intensity.

The direct precision methods show no clear monotonic trend on Lorenz~96 but improve steadily on cosmological data (Table~\ref{tab:cosmo_enkf}). The difference is structural: cosmological data has approximately diagonal precision, so the diagonal targets capture its dominant structure, whereas Lorenz~96 has dense off-diagonal precision that none of the generic targets approximate well. The dual-domain experiments in Section~\ref{sec:cosmo_validation} confirm this directly.

When domain-specific targets are unavailable, covariance-domain approaches like Ledoit-Wolf are the safer choice.

\subsection{Educational Aspects}

TEDA's modular design lets students swap precision matrix estimators by changing a single parameter while keeping the dynamical model, ensemble size, and observation network fixed. Variable names in the code match the mathematical notation in this paper, and the experimental results themselves illustrate how assumptions and domain knowledge shape the performance of otherwise elegant estimators.

\subsection{Estimation Quality vs.\ Filtering Performance}
\label{sec:estimation_vs_filtering}

The most surprising result is the disconnect between standalone estimation quality and filtering performance. Ledoit-Wolf produces the worst standalone precision estimates on cosmological data (Table~\ref{tab:cosmo_direct}) yet the best EnKF RMSE (Table~\ref{tab:cosmo_enkf}). The explanation is sequential accumulation: the precision estimator is applied at every assimilation cycle, and step-to-step variability compounds across 100 updates. Ledoit-Wolf's standard deviation across draws is $\pm 0.01$ (Part~A)---an order of magnitude below the direct precision methods ($\pm 0.13$--$0.14$). A slightly biased but stable estimator beats a more accurate but variable one when errors accumulate over many cycles. For sequential data assimilation, variance reduction matters at least as much as bias reduction.

\subsection{Domain Dependence: Lorenz~96 vs.\ Cosmological Data}
\label{sec:domain_dependence}

The dual-domain comparison confirms that the poor Lorenz~96 results for direct precision shrinkage reflect target-truth mismatch, not a flaw in the methods. On Lorenz~96, diagonal targets produce erratic RMSE ($0.43$--$2.00 \times 10^{-2}$). On the near-diagonal cosmological data, the same methods achieve stable performance ($1.73$--$2.39 \times 10^{-2}$). The Bodnar framework works as designed when the target captures the dominant structure of the true precision matrix.

