\section{Introduction}
\label{sec:intro}

Ensemble Kalman filters require the inverse of the background error covariance matrix, but in most operational settings the ensemble is far too small to estimate that matrix reliably. When the state dimension $n$ exceeds the ensemble size $N_e$, the sample covariance is rank-deficient and cannot be inverted at all~\cite{houtekamer1998data}; even when $N_e > n$, sampling noise introduces spurious correlations that contaminate the precision matrix~\cite{anderson2001ensemble}. The same problem appears in cosmological data analysis, where the number of mock catalogs available to estimate power spectrum covariances is often comparable to the data dimension~\cite{hamilton2006measuring}.

Shrinkage estimation offers a practical remedy: combine the noisy sample estimate with a structured, low-variance target matrix, accepting a small bias in exchange for substantial variance reduction. Most of the established literature operates in covariance space. Ledoit and Wolf~\cite{ledoit2004well} derived a closed-form optimal intensity for shrinking toward a scaled identity; Pope and Szapudi~\cite{pope2008shrinkage} introduced shrinkage to cosmology; Chen et al.~\cite{chen2010shrinkage} refined it with an oracle-approximating estimator. Bodnar et al.~\cite{bodnar2016direct} took a different route, shrinking the precision matrix directly rather than first shrinking the covariance and then inverting. Looijmans et al.~\cite{looijmans2024comparison} applied this direct precision framework to BOSS DR12 mock catalogs, showing that a domain-specific (cosmological) target outperforms generic alternatives in a standalone setting.

None of the existing approaches is without practical drawbacks. Covariance-space shrinkage minimizes $\mathbb{E}\|\hat{\boldsymbol{\Sigma}} - \boldsymbol{\Sigma}\|_F^2$, not the precision-space loss that governs the EnKF update; the inverse of a covariance-optimal estimator amplifies errors in small eigenvalues, which are exactly the directions where regularization is needed. Target-free methods such as NERCOME~\cite{lam2016nonparametric} buy robustness at the cost of repeated bootstrap eigendecompositions that are prohibitive inside a sequential filter. Sparse-precision approaches (graphical lasso~\cite{friedman2008sparse}, neighborhood selection~\cite{meinshausen2006high}) require a sparsity assumption that is not satisfied by either Lorenz-like dense precision or nearly-diagonal power-spectrum precision. Direct precision shrinkage~\cite{bodnar2016direct} removes the inversion bias but is only defined for invertible sample covariances and its published target menu is limited to identity, eigenvalue-based, and domain-specific (cosmological) options. Localization~\cite{gaspari1999construction} side-steps the problem by imposing sparsity but destroys long-range correlations and invalidates the Wishart assumption underlying the shrinkage formulas.

Two concrete gaps remain. First, Looijmans et al.~\cite{looijmans2024comparison} applied these estimators to standalone precision estimation, not inside a sequential filter. The EnKF recomputes the precision matrix at every assimilation cycle, and errors in the precision estimate feed back through the forecast model into subsequent cycles. It is unclear whether an estimator that performs well in a one-shot setting will also perform well when applied repeatedly across hundreds of cycles. Second, the direct-precision target menu lacks a middle ground between the plain identity (ignores variable-specific scales) and the cosmological target (requires domain knowledge), leaving practitioners without a domain-agnostic but scale-aware option.

\paragraph{Contributions.}
We address both gaps and make the following contributions.
(i)~We embed the Bodnar et al.~\cite{bodnar2016direct} direct precision shrinkage framework inside the dual EnKF formulation, recomputing the shrinkage coefficients at each assimilation cycle from the current ensemble, and integrate six estimators into the TEDA toolbox~\cite{nino2022teda}.
(ii)~We propose a \emph{scaled identity target} $\boldsymbol{\Pi}_0 = \mathrm{diag}(1/(2\sigma_i^2))$ derived from the empirical per-variable variances of the sample covariance, providing a domain-agnostic alternative to the cosmological target.
(iii)~We benchmark the six estimators on two systems with opposite precision structure---Lorenz~96~\cite{lorenz1996predictability} ($n = 40$, dense off-diagonal precision) and BOSS DR12 mocks~\cite{looijmans2024comparison} ($d = 18$, near-diagonal precision)---and show that (a)~target-truth alignment determines direct precision shrinkage performance, with the methods failing on Lorenz~96 and succeeding on cosmological data, and (b)~good standalone estimation does not imply good filtering, because cycle-to-cycle variance compounds over repeated assimilation updates. Ledoit--Wolf's low-variance estimator achieves the lowest EnKF RMSE on cosmological data despite having the worst standalone Frobenius loss.

The remainder of the paper is organised as follows. Section~\ref{sec:preliminaries} reviews the EnKF, the Wishart distribution of sample covariances, and the main families of shrinkage estimators. Section~\ref{sec:method} presents the direct precision shrinkage framework, the four targets we evaluate, and the proposed scaled identity target. Section~\ref{sec:benchmark} describes the dual-domain benchmark and reports results. Section~\ref{sec:discussion} analyses the sequential accumulation effect and its implications for estimator design. Section~\ref{sec:conclusions} concludes.
