\section{Preliminaries}
\label{sec:preliminaries}

\subsection{The Ensemble Kalman Filter}
\label{sec:enkf}

The Ensemble Kalman Filter~\cite{evensen2003ensemble} represents the background state with an ensemble of $N_e$ state vectors $\{\mathbf{x}_i^b\}_{i=1}^{N_e}$, each $\mathbf{x}_i^b \in \mathbb{R}^n$, whose mean $\bar{\mathbf{x}}^b = N_e^{-1}\sum_i \mathbf{x}_i^b$ is the forecast estimate. Writing the deviation matrix $\Delta\mathbf{X}^b = \mathbf{X}^b - \bar{\mathbf{x}}^b\mathbf{1}^T \in \mathbb{R}^{n \times N_e}$, the sample background covariance is
\begin{equation}
\label{eq:sample_cov}
\mathbf{S} \;=\; \mathbf{P}^b \;=\; \frac{1}{N_e - 1}\,\Delta\mathbf{X}^b (\Delta\mathbf{X}^b)^T \;\in\; \mathbb{R}^{n \times n}.
\end{equation}
The stochastic analysis solves $[\mathbf{H}\mathbf{P}^b\mathbf{H}^T + \mathbf{R}]\,\Delta\mathbf{X}^a = \mathbf{D}^s$ for the deviation of the analysis ensemble, with $\mathbf{D}^s$ the matrix of perturbed innovations. The \emph{dual} formulation~\cite{nino2018ensemble} uses the background \emph{precision} directly,
\begin{equation}
\label{eq:dual_enkf}
\bigl[(\mathbf{P}^b)^{-1} + \mathbf{H}^T\mathbf{R}^{-1}\mathbf{H}\bigr]\,\Delta\mathbf{X}^a \;=\; \mathbf{H}^T\mathbf{R}^{-1}\mathbf{D}^s,
\end{equation}
and is the entry point for this work: plugging a shrinkage-based precision estimator into~\eqref{eq:dual_enkf} gives a drop-in replacement for the sample precision $\mathbf{S}^{-1}$ at each assimilation cycle.

\subsection{The Wishart Distribution and Sample Precision}
\label{sec:wishart}

If the ensemble members are drawn from a Gaussian with true covariance $\boldsymbol{\Sigma}$, then $(N_e - 1)\mathbf{S} \sim \mathcal{W}_n(N_e - 1, \boldsymbol{\Sigma})$ is Wishart-distributed. Two properties matter for what follows. First, the eigenvalues of $\mathbf{S}$ are systematically spread relative to $\boldsymbol{\Sigma}$: the largest are overestimated, the smallest underestimated, and the bias grows with $n/N_e$. Inverting $\mathbf{S}$ turns the overestimated small eigenvalues into underestimated large eigenvalues of $\mathbf{S}^{-1}$---and vice versa---so the sample precision inherits the spreading with amplified noise on the small-$N_e$ end. Second, when $N_e - 1 > n + 1$, $\mathbb{E}[\mathbf{S}^{-1}] = \frac{N_e-1}{N_e-n-2}\,\boldsymbol{\Sigma}^{-1}$, so $\mathbf{S}^{-1}$ is a biased estimator. The Hartlap correction $\gamma = (N_e - n - 2)/(N_e - 1)$~\cite{hartlap2007your} removes this bias in standalone use, but the Bodnar et al.~\cite{bodnar2016direct} direct-precision shrinkage coefficients already account for it through the $(1 - n/N_e)$ term, so it should \emph{not} be pre-applied when using those formulas.

\subsection{Covariance Matrix Shrinkage}
\label{sec:cov_estimation}

Covariance shrinkage blends $\mathbf{S}$ with a structured target $\mathbf{T}$,
\begin{equation}
\label{eq:shrinkage_cov}
\hat{\mathbf{P}} \;=\; (1 - \lambda)\,\mathbf{S} \,+\, \lambda\,\mathbf{T}, \qquad \lambda \in [0,1],
\end{equation}
and chooses $\lambda$ to minimize $\mathbb{E}\|\hat{\mathbf{P}} - \boldsymbol{\Sigma}\|_F^2$. Ledoit and Wolf~\cite{ledoit2004well} derived a closed-form data-driven intensity for $\mathbf{T} = \mu\mathbf{I}$, $\mu = \mathrm{tr}(\mathbf{S})/n$, that is asymptotically optimal within this class and has become a standard reference. Chen et al.~\cite{chen2010shrinkage} refined the result with an oracle-approximating variant that is particularly effective in the small-$N_e$ regime. Both estimators produce well-conditioned covariances but still require an explicit inversion to obtain a precision matrix, and that inversion amplifies errors along the smallest sample eigenvalues---precisely the directions where shrinkage was supposed to regularize.

\paragraph{NERCOME.}
NERCOME~\cite{lam2016nonparametric,joachimi2017non} avoids the target-choice problem by splitting the ensemble into disjoint subsets, diagonalizing the first subset's sample covariance, rotating the second subset into that eigenbasis, and keeping only the diagonal. Averaging over $B$ random splits debiases the eigenvalue estimates nonparametrically. The price is $\mathcal{O}(B)$ eigendecompositions per estimate, which becomes prohibitive inside a sequential filter.

\subsection{Precision Matrix Shrinkage}
\label{sec:precision_estimation}

Rather than shrinking $\mathbf{S}$ and inverting, Bodnar et al.~\cite{bodnar2016direct} proposed shrinking the precision matrix \emph{directly}. The estimator combines the sample precision $\hat{\boldsymbol{\Pi}}_S = \mathbf{S}^{-1}$ with a structured precision target $\boldsymbol{\Pi}_0$,
\begin{equation}
\label{eq:direct_precision}
\boldsymbol{\Pi}_{\mathrm{LS}} \;=\; \alpha^*\,\hat{\boldsymbol{\Pi}}_S \,+\, \beta^*\,\boldsymbol{\Pi}_0,
\end{equation}
with analytically derived coefficients
\begin{align}
\label{eq:alpha_star}
\alpha^* &= 1 - \frac{n}{N_e} - \frac{N_e^{-1}\,\|\hat{\boldsymbol{\Pi}}_S\|_{\mathrm{tr}}^2\,\|\boldsymbol{\Pi}_0\|_F^2}{\|\hat{\boldsymbol{\Pi}}_S\|_F^2\,\|\boldsymbol{\Pi}_0\|_F^2 - \bigl(\mathrm{tr}(\hat{\boldsymbol{\Pi}}_S \boldsymbol{\Pi}_0)\bigr)^2}, \\
\label{eq:beta_star}
\beta^* &= \left(1 - \frac{n}{N_e} - \alpha^*\right)\frac{\mathrm{tr}(\hat{\boldsymbol{\Pi}}_S \boldsymbol{\Pi}_0)}{\|\boldsymbol{\Pi}_0\|_F^2},
\end{align}
that minimize the expected quadratic loss $\mathbb{E}\|\boldsymbol{\Pi}_{\mathrm{LS}} - \boldsymbol{\Sigma}^{-1}\|_F^2$ under the Wishart assumption. The denominator in~\eqref{eq:alpha_star} is non-negative by Cauchy--Schwarz and vanishes only when $\hat{\boldsymbol{\Pi}}_S \propto \boldsymbol{\Pi}_0$; when the target is a close match, the correction term drives $\alpha^*$ down and $\beta^*$ up (more weight to the target), and vice versa. In practice the unclamped coefficients can leave $[0,1]$, in which case Bodnar et al.~\cite{bodnar2016direct} clamp them before use.

\paragraph{Cosmological precision structure.}
For power-spectrum measurements of a Gaussian random field~\cite{hamilton2006measuring},
\begin{equation}
\label{eq:cosmo_cov}
\mathrm{Cov}(\hat{C}_{\ell_i}, \hat{C}_{\ell_j}) \;=\; \frac{2}{(2\ell_i+1)f_{\mathrm{sky}}}\!\left(C_{\ell_i} + \frac{1}{n_g}\right)^{\!2}\!\delta_{\ell_i\ell_j},
\end{equation}
is diagonal by mode independence. Pope and Szapudi~\cite{pope2008shrinkage}, as adopted by Looijmans et al.~\cite{looijmans2024comparison}, use the leading-order cosmic-variance part as a precision target via $T^{(2)}_{ii} = (2/N_{\ell_i})\,C_{\ell_i}^2$, with $N_{\ell_i} = (2\ell_i+1)f_{\mathrm{sky}}$, and $\boldsymbol{\Pi}_0^{(2)} = (\mathbf{T}^{(2)})^{-1}$.

\subsection{Localization and Modified Cholesky}
\label{sec:localization}

In the data-assimilation community, the dominant remedy for rank-deficient ensemble covariances is \emph{localization}: element-wise multiplication of $\mathbf{S}$ by a compactly supported correlation function~\cite{gaspari1999construction,houtekamer1998data} that zeros out long-range correlations. The \emph{modified Cholesky decomposition}~\cite{nino2018ensemble} is a banded variant that factors $(\mathbf{P}^b)^{-1} \approx \mathbf{L}^{-T}\mathbf{D}^{-1}\mathbf{L}^{-1}$ with $\mathbf{L}$ lower triangular and $\mathbf{D}$ diagonal, estimating each row of $\mathbf{L}$ by regressing a state variable on its neighbors within a bandwidth $r$. The cost is $\mathcal{O}(n\,r\,N_e)$ per cycle, linear in $n$ for fixed $r$, which is the only scaling in our comparison that survives operational state dimensions. Localization and shrinkage are complementary: localization exploits spatial structure when it is known, shrinkage exploits similarity to a target when no such structure is available.

\subsection{Critical Comparison of Existing Approaches}
\label{sec:critical}

The estimators reviewed above address the same problem from complementary angles, but none is uniformly preferable, and their limitations jointly define the gap this paper addresses.

\emph{Shrinkage in covariance space is not shrinkage in precision space.} Ledoit and Wolf~\cite{ledoit2004well} minimizes $\mathbb{E}\|\hat{\boldsymbol{\Sigma}} - \boldsymbol{\Sigma}\|_F^2$, not the precision loss that governs the EnKF update; the inverse of a covariance-optimal estimator amplifies errors in small eigenvalues, echoing the classical ridge-regression observation that regularizing a matrix and its inverse are different problems~\cite{hoerl1970ridge}. Cosmology has explored surrogate precision priors directly~\cite{paz2015improving,taylor2014estimating,sellentin2016parameter,dodelson2013effect}, but without porting them to sequential filtering.

\emph{Target-free methods pay for their generality.} NERCOME removes the target-choice problem at the cost of $\mathcal{O}(B)$ eigendecompositions per cycle. If a reasonable target exists, a single-pass linear shrinkage is cheaper \emph{and} lower-variance; if no target is available, NERCOME's cost becomes prohibitive inside the EnKF loop (see Section~\ref{sec:results}).

\emph{Diagonal targets assume the problem away.} The identity, eigenvalue, and cosmological targets available in the direct-precision framework are all diagonal. This is appropriate for data whose true covariance is near-diagonal by construction (Gaussian random fields at independent Fourier modes) but not for dynamical systems with spatial coupling, where the true precision is \emph{dense} and off-diagonal. Shrinking toward a diagonal target then forces the Bodnar formulas to trade off two wrong ingredients, producing miscalibrated coefficients that we see directly in Section~\ref{sec:results}.

\emph{Standalone optimality is not sequential optimality.} Every estimator above is designed for a single-shot loss. Inside an EnKF, the precision is recomputed at every cycle, and errors accumulate over hundreds of updates. A low-bias, high-variance estimator that dominates in a single-shot comparison can easily be outperformed over 100 cycles by a higher-bias, lower-variance alternative; this effect is the central empirical finding of this paper.
