\section{Conclusions}
\label{sec:conclusions}

A good precision matrix estimator and a good filtering estimator are not the same thing. On cosmological data, the Bodnar et al.~\cite{bodnar2016direct} methods achieve Frobenius losses half those of Ledoit-Wolf, yet Ledoit-Wolf yields the lowest EnKF RMSE. The reason is stability: a slightly biased estimator that behaves consistently from cycle to cycle accumulates less error over 100 sequential updates than one that is accurate on average but noisy. Without domain-specific prior knowledge, Ledoit-Wolf is the safest default among the methods we tested.

Direct precision shrinkage remains the better choice when the target matches the true precision structure. Our cosmological experiments confirm Looijmans et al.~\cite{looijmans2024comparison}: with near-diagonal data, these methods outperform all alternatives in standalone estimation. Their failure on Lorenz~96 is a consequence of applying diagonal targets to a system whose precision matrix has dense off-diagonal structure, not a deficiency of the framework itself.

Our implementation does not combine shrinkage with spatial localization~\cite{gaspari1999construction,nino2018ensemble}, which could yield further gains. The scaled identity target uses an ad-hoc factor of~2 without theoretical justification beyond analogy to the cosmological $2/N_\ell$ term. The AR(1) model in Part~B preserves the covariance structure of the BOSS mock catalogs but is far simpler than a realistic cosmological simulation. Future work should address the interaction between shrinkage and localization, test on higher-dimensional systems such as the quasi-geostrophic model, and explore adaptive target selection based on observed covariance structure.

All implementations are publicly available in TEDA, with code structured to mirror the algorithms in this paper.

\subsubsection{Acknowledgements.} This work was supported by the Department of Computer Science and Statistics at Universidad del Norte. The authors thank the TEDA development team for the foundational framework.
